\paperTitle

### 2014-04-28 doc/icml2014/background.tex
changes
Fixed several minor changes brought up during rebuttals
Updated paper (draft to send to David Sontag)

% PL: don't know what this means
%We assume that the model $\sG$ is parameterized at every clique $\sC$.
If $\sG$ is directed, the model parameters correspond to
the conditional probability tables for each clique $(\Pa(a), a) \in \sG$.
If $\sG$ is undirected, we assume a log-linear parametrization of each clique $\sC \in \sG$.

@@

% Maximum likelihood
The standard estimation procedure is maximum (marginal) likelihood,
  \begin{align*}
    L_\text{unsup}(\theta) &\eqdef \max_{\theta} \sum_{\vx \in \sD} \log p_\theta(\vx) \\
        &=      \max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} p_\theta(\vx,\vh).
  \end{align*}
  Maximum likelihood is statistically efficient,
  but in general computationally intractable
  because marginalizing over hidden variables $\vh$ yields a non-convex objective.
In practice, one uses local optimization procedures (e.g., EM
  or L-BFGS) on the marginal likelihood, which can get stuck in local
  optima.
We will later return to likelihoods, but let us first
describe a method of moments approach for parameter estimation.
The starting point is moments of the marginal distribution $\BP(\vx)$,
and a bit of notation is necessary.

@@

We use the notation $[\cdot]$ to indicate indexing; for example $M[i]$
  is the $i$-th row of a matrix $M$ and $M[i,j]$ is the $i,j$-th element
  of $M$.
For a tensor $T \in \Re^{d \times \ldots \times d}$ and a vector $\vec
  i = (i_1, \cdots, i_\ell)$, define the projection $T[\vec i] = T_{i_1 \ldots i_\ell}$.
Similarly, for a set $\sC$ of size $k$ and vector $\vec x \in \Re^d$, we
  will use the notation $\vec x_\sC$ to get a vector $\Re^k$ which selects
  the appropriate elements of $\vec x$.

@@

We consider up to third-order moments:
\begin{align*}
  \mO_i &\eqdef \Pr(x_i), &
  \mO_{ij} &\eqdef \Pr(x_i, x_j), &
  \mO_{ijk} &\eqdef \Pr(x_i, x_j, x_k).
\end{align*}
We will view $\mO_i, \mO_{ij}, \mO_{ijk}$ as tensors of
  orders $1, 2, 3$ in $\Re^d, \Re^{d\times d}, \Re^{d \times d \times d}$.
Next, define the \emph{hidden marginals}:
\begin{align*}
  \mH_i &\eqdef \Pr(h_i), &
  \mH_{ij} &\eqdef \Pr(h_i, h_j), &
  \mH_{ijk} &\eqdef \Pr(h_i, h_j, h_k).
\end{align*}
These are tensors of
  orders $1, 2, 3$ in $\Re^k, \Re^{k\times k}, \Re^{k \times k \times k}$.

@@

\citet{anandkumar12moments} presented a consistent estimator
  using the method of moments to learn the parameters of a three-view
  mixture model (\figureref{three-view}), an important case of directed
  graphical models. %, that sidesteps the local optima problem altogether. 
In that model, we have three observed variables $x_1, x_2, x_3$
  that are independent conditioned on a hidden variable $h_1$.  
The parameters of this model are the mixture probabilities $\pi\oft 1 \eqdef
  \Pr(h_1) \in \Delta_{k-1}$ and the conditional moments $\mOpp{i}{1} \eqdef
  \Pr(x_i \given h_1) \in \Re^{d \times k}$ for $i \in \{1,2,3\}$.

@@

\citet{anandkumar13tensor} presented a robust tensor power method, which
  we call $\TensorFactorize$, which converts estimates of the observed moments 
  into estimates of the parameters, $\pi\oft 1, \mOpp{1}{1},
  \mOpp{2}{1},\mOpp{3}{1}$ (up to permutation of the columns), provided
  the following assumption holds:

@@

In brief, the algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^T \mO_{12} W = I_{k \times k}$,
  and uses it to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$.
  $\mOpp{1}{1}$ and $\mOpp{2}{1}$ can then be recovered with a few matrix operations.

>>>>

% PL: don't know what this means
%We assume that the model $\sG$ is parametrized at every clique $\sC$.
If $\sG$ is directed, the model parameters correspond to
the conditional probability tables for each clique $(\Pa(a), a) \in \sG$.
If $\sG$ is undirected, we assume a log-linear parametrization of each clique $\sC \in \sG$.

@@

% Maximum likelihood
The standard estimation procedure is maximum (marginal) likelihood,
  \begin{align*}
    L_\text{unsup}(\theta) &\eqdef \sum_{\vx \in \sD} \log p_\theta(\vx)
        =  \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} p_\theta(\vx,\vh).
  \end{align*}
  Maximum likelihood is statistically efficient,
  but in general computationally intractable
  because marginalizing over hidden variables $\vh$ yields a non-convex objective.
In practice, one uses local optimization procedures (e.g., EM
  or L-BFGS) on the marginal likelihood, which can get stuck in local
  optima.
We will later return to likelihoods, but let us first
describe a method of moments approach for parameter estimation.
The starting point is moments of the marginal distribution $\BP(\vx)$,
and a bit of notation is necessary.

@@

We use the notation $[\cdot]$ to indicate indexing; for example $M[i]$
  is the $i$-th row of a matrix $M$ and $M[i,j]$ is the $i,j$-th element
  of $M$.
For a tensor $T \in \Re^{d \times \ldots \times d}$ and a vector $\vec
  i = (i_1, \cdots, i_\ell)$, define the projection $T[\vec i] = T_{i_1 \ldots i_\ell}$.
Similarly, for a set $\sC$ of size $k$ and vector $\vec x \in \Re^d$, we
  will use the notation $\vec x_\sC$ to get a vector in $\Re^k$ which selects
  the appropriate elements of $\vec x$.

@@

We consider up to third-order moments:
\begin{align*}
  \mO_i &\eqdef \Pr(x_i), &
  \mO_{ij} &\eqdef \Pr(x_i, x_j), &
  \mO_{ijk} &\eqdef \Pr(x_i, x_j, x_k).
\end{align*}
We will view $\mO_i, \mO_{ij}, \mO_{ijk}$ as tensors of
  orders $1, 2, 3$ in $\Re^d, \Re^{d\times d}, \Re^{d \times d \times d}$.
Next, we define the \emph{hidden marginals}:
\begin{align*}
  \mH_i &\eqdef \Pr(h_i), &
  \mH_{ij} &\eqdef \Pr(h_i, h_j), &
  \mH_{ijk} &\eqdef \Pr(h_i, h_j, h_k).
\end{align*}
These are tensors of
  orders $1, 2, 3$ in $\Re^k, \Re^{k\times k}, \Re^{k \times k \times k}$.

@@

\citet{anandkumar12moments} presented a consistent estimator
  using the method of moments to learn the parameters of a three-view
  mixture model (\figureref{three-view}), an important case of directed
  graphical models. %, that sidesteps the local optima problem altogether. 
In that model, we have three observed variables $x_1, x_2, x_3$
  that are independent conditioned on a hidden variable $h_1$.  
The parameters of this model are the mixture probabilities $\pi\oft 1 \eqdef
  \Pr(h_1) \in \Re^k$ and the conditional moments $\mOpp{i}{1} \eqdef
  \Pr(x_i \given h_1) \in \Re^{d \times k}$ for $i \in \{1,2,3\}$.

@@

\citet{anandkumar13tensor} presented a robust tensor power method, which
  we call $\TensorFactorize$, which transforms estimates of the observed moments 
  into estimates of the parameters, $\pi\oft 1, \mOpp{1}{1},
  \mOpp{2}{1},\mOpp{3}{1}$ (up to permutation of the columns), provided
  the following assumption holds:

@@

In brief, the algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^\top \mO_{12} W = I_{k \times k}$,
  and uses it to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$.
  $\mOpp{1}{1}$ and $\mOpp{2}{1}$ can then be recovered with a few matrix operations.
